A Survey of Possible Exoskeleton Control Architectures and Algorithms\\ (Draft 2.0)

### 2015-09-20 control.tex
split into 3 papers

\title{A Survey of Exoskeleton Control (Draft 1.1)}

@@

\section{What is this paper about?}

This paper surveys state of the art exoskeleton technology with an
emphasis on control. The focus is on exoskeleton technologies that allow a
highly trained and top percentile athletic 
operator to carry a payload that weighs approximately the same amount
as the operator. We envisage these types of exoskeletons to be useful
in carrying protective and safety equipment for SWAT teams, police,
firefighters, and soldiers.

@@

We do not survey exoskeleton technology for human arms. 
We treat the torso and helmet of the exoskeleton as the payload,
and focus on a lower body exoskeleton to support any payloads that
are on the torso or head.
We do not cover global energy storage such as batteries or liquid fuel.

\section{What would it be like for the operator?}

@@

Bypass valves and clutches allow the operator to physically ``take over''
and push the relatively lightweight leg mounted exoskeleton around.

``Passive'' degrees of freedom can have physical or ``virtual'' springs
applied to assist in support (all abduct/adduct DOF, rotational DOF).
These aids can be clutched out or turned off when necessary.
``Virtual'' sprintgs can be modulated.

@@

Can we build an exoskeleton that allows the operator/exoskeleton team
to be effective? The operator needs to learn to ``fly'' the suit,
just as operators learn to operate parachutes,
wing suits, diving equipment, high and low temperature protective
suits, firefighting equipment, and high altitude flight suits.

@@

Useful paper: \url{http://summerschool.stiff-project.org/fileadmin/pdf/1804_C19.pdf}

Variants of nonlinear feedback control such as feedback linearization or
sliding mode control are largely ignored. Once the decision to use feedback control
based on a set of observable quantities and with particular outputs is made,
one can try out the various linear and nonlinear feedback control paradigms
to see what works well.
Variants of function approximation methods such as lookup tables, fuzzy logic,
sigmoidal neural networks, radial basis functions, and locally weighted regression
are also largely ignored. Once the decision to use a function approximator
and what the inputs and outputs has been made, one can try out the various
approaches to see what works well.
Same for optimization methods.
Same for constraint enforcement (such as avoiding self-collisions)
in either optimization or feedback control (barrier
Lyapunov functions (BLF)~\cite{IEEE06911561}).

Stability proofs of any of these methods should be viewed extremely skeptically due
to the unmodeled operator-exoskeleton and exoskeleton-world contact dynamics, 
actuator unmodeled dynamics, joint play and structural deformation, controller
time delays which are typically ignored, especially in proofs involving
passivity arguments or Lyapunov functions.

The dynamics of a revolute exoskeleton are:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \mJ_w^{\tr}(\vq) \vf_w = \vtau + \mJ_o^{\tr}(\vq) \vf_o
\label{eq:dynamics}
\end{equation}
where $\vq$ are the joint angles and position and velocity of the ``root'' of
the exoskeleton, $\mM$ is the inertia matrix, $\mC()$ are the Coriolis and
centripetal forces, $\mG()$ are the gravitational forces, $\mF$ are the
friction forces, $\mJ_w^{\tr}(\vq) \vf_w$ (Jacobian matrix multiplied with the
contact forces)
are the exoskeleton-world
contact forces 
expressed as exoskeleton joint torques, 
$\vtau$ are the exoskeleton
joint torques,
and $\mJ_o^{\tr}(\vq) \vf_o$
are the operator-exoskelton forces expressed
as exoskeleton joint torques,
$\vtau$ is augmented with zeros corresponding to the dimensions of $\dot{\vq}$
that correspond to root velocities, since those dimensions are not actuated.

To understand these equations better, it is useful to have all Jacobians equal
to the identity matrix. This means the operator is directly applying torques
at the exoskeleton joints ($\vtau_o$), and so is the world ($\vtau_w$). In this case:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \vtau_w = \vtau + \vtau_o
\end{equation}

Note that we are ignoring the operator dynamics and operator muscle and tissue
forces resulting in operator joint torques. We will add these in later (???).
We are also ignoring actuator dynamics, which we can handle by augmenting 
the state vector.

\subsection{No control}

Rely on counterbalancing, mechanical springs, air springs, etc.
to reduce forces. 
Operator relies on mechanical backdrivability to drive
exokeleton to do task.

$\vtau$ is generated by passive devices such as springs or dampers.

\subsection{Closed loop behavior selection, open loop behavior
execution}

Use joint/Cartesian position and velocity control of exoskeleton 
with pre-generated references or selected targets and automatically generated trajectories to move to targets or perform tasks.
Getting behavior selection is very important, operator may do it with
special interface (keyboard, game controller, ...).
Operator relies on mechanical backdrivability to correct exoskeleton behavior.

Recognizing/estimating phase goes here.

Phase Triggered References could go here.

@@

central pattern generator

iterative learning control to acquire $\vtau_{ff}(t)$ for desired trajectories.

$\vtau = \vtau_{ff}(t)$ where $\vtau_{ff}$ is a time
dependent feedforward torque vector.

\subsection{Active gravity compensation}

Similar to counterbalancing or physical gravity compensation, except done
with exoskeleton actuators to cancel exoskeleton weight (but not inertia).
A simple formulation is:
\begin{equation}
\vtau = \mJ^{\tr} (m\vg)
\end{equation}
where $\vtau$ are the exoskeleton joint torques, 
$\mJ$ is a Jacobian matrix, $m$ is the weight of the exoskeleton,
and $\vg$ is the gravity vector. 
The operator relies on mechanical backdrivability to drive
the exokeleton to do tasks.

Sarcos Primus Humanoid (ATR) video:\\
\url{https://www.youtube.com/watch?v=KNxxLm4sPys}

$\vtau = \hat{\mG}(\hat{\vq})$ 
are the estimated gravitational torques at the estimated exoskeleton configuration.

\subsection{Friction compensation}

$\vtau = \hat{\mF}(\hat{\vq},\hat{\dot{\vq}})$ 
are the estimated friction torques at the estimated exoskeleton configuration
and velocity.
Note that these may be load dependent, and may require additional state to
correctly handle stiction and hysteresis effects.

\subsection{Online inverse dynamics control}

The goal here is to map from desired motion (in this case desired
accelerations) to exoskeleton actuator commands (and desired operator
and world contact forces).
The exoskeleton actuator torque 
attempts to match the torques caused by inertial, Coriolis, centripetal,
gravitational, frictional, and exoskeleton-world contact forces.
It relies on estimated dynamic models and their parameters.
If there are actuator dynamics or other dynamics, we can include those
dynamics as well.
\begin{equation}
\vtau = \vtau_{\invdyn} 
= \hat{\mM}(\hat{\vq}) \ddot{\vq}_d
+ \hat{\mC}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mG}(\hat{\vq})
+ \hat{\mF}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mJ}_w^{\tr}(\hat{\vq}) \hat{\vf}_w
\end{equation}
If the models and estimated state (position and velocity) are perfect, we get
perfect cancellation of most of the dynamics:
\begin{equation}
\mM(\vq) (\ddot{\vq} - \ddot{\vq}_d) = \mJ_o^{\tr}(\vq) \vf_o
\end{equation}
If $\ddot{\vq}_d$ is set to zero, the operator directly drives the acceleration
of the exoskeleton with an effective inertia of $(\mM^{-1}(\vq) \mJ_o^{\tr}(\vq))^{-1}$
\begin{equation}
\ddot{\vq} = \mM^{-1}(\vq) \mJ_o^{\tr}(\vq) \vf_o
\label{eq:cancel}
\end{equation}
$\ddot{\vq}_d$ can be used to drive the exoskeleton along trajectories, and
the operator can directly modify the trajectory.

@@

Online inverse dynamics control is a generalization of active gravity compensation to include
inertial, Coriolis, and centripetal forces (forces due to acceleration and
velocity).
If frictional forces are modeled they can also be canceled.
Online inverse dynamics can be added to open loop behavior execution
and many other types of control such as operator force
feedback control, impedance control, and admittance control.
Computed torque control and feedback linearization are forms of online
inverse dynamics.

@@

\subsection{Impedance Control}

@@

\subsection{Force feedback and get out of the way control (inverse dynamics version)}

@@

\subsection{Mapping External Forces To Operator Forces}

\subsection{Creating An External Impedance}

\subsection{Mapping External Impedances To Operator Impedance}

\subsection{Constrained Inverse Dynamics: Quadratic Programming}

The dominant paradigm in force controlled (typically
hydraulic) full size humanoid robots is to handle constraints in inverse dynamics
such as torque and center of pressure limits using online quadratic programming
to solve the inverse dynamics equations.

Quadratic programming~\cite{Wikipedia} solves the following problem:
Minimize
\begin{equation}
\vx^{\tr} \mQ \vx + c^{\tr} \vx
\end{equation}
with inequality constraints
\begin{equation}
\mA_1 \vx \le \vb_1
\end{equation}
and equality constraints
\begin{equation}
\mA_2 \vx = \vb_2
\end{equation}
The optimization variable $\vx$ includes accelerations $\ddot{\vq}$, joint
torques $\vtau$, and contact forces $\vf$.
For robots the discrepancy between actual and desired accelerations, joint
torques, and contact forces are all minimized:
\begin{equation}
C_r(\vx) = (\vq - \vq_d)^{\tr} \mW_{\vq} (\vq - \vq_d) + \vtau^{\tr} \mW_{\vtau} \vtau
+ \vf^{\tr} \mW_{\vf} \vf
\end{equation}
The dynamics (equation~\ref{eq:dynamics}) are enforced as an equality constraint.
Torque limits, friction limits, center of pressure location constraints,
and unidirectional contact constraints are all
enforced as inequality constraints.
These optimizations were performed every 1-2ms by various teams to control the
Atlas robots in the DARPA Robotics Challenge.

@@

\subsection{Operator-Exoskeleton Force Control using position controlled actuation}

@@

\subsubsection{Get out of the way control}

@@

\subsubsection{Operator force feedback}

@@

\subsubsection{Integral admittance control}

@@

We are skeptical that acceleration feedback (even low pass filtered acceleration
estimated by a Kalman filter) will
work well on real systems. Impacts, shock waves, and the fact that neither
the operator's body parts or the exoskeletons parts are rigid bodies and the
joints are not well defined suggest this approach would be fragile to modeling
error and unmodeled dynamics.

@@

@@

Sensitivity Amplification Control (SAC) used to control BLEEX
is a dynamic cancellation technique (a.k.a.
inverse dynamics, computed torque, or feedback linearization) that also tries
to adjust the apparent inertia of the exoskeleton (equation~\ref{eq:change-inertia1})
using acceleration feedback. It is not clearly stated but it appears the
acceleration is the result of double differentiating position, as there do not
appear to be velocity sensors on BLEEX.

stance: virtual joint torque control method Kazerooni et al.(2005)
When contact location between the wearer and the
exoskeleton is not fixed and difficult to estimate, this method
has been shown to be an effective method to generate the
locomotion for an exoskeleton robot.[3],[4],[21]
\begin{verbatim}
H. Kazerooni , Z. Racine , L. Huang and R. Steger ”On the control of
the Berkeley lower extremity exoskeleton (BLEEX),” Proc. IEEE Int.
Conf. Robot. Autom., pp. 4364-4371, 2005.
Racine Jean-Louis Charles ”Control of a Lower Extrmity Exoskeleton
for Human Performance Amplification,” University of California,2003.
582
Xiuxia Yang, Hongchao Zhao, Yi Zhang, Xiaowei Liu ”Carrying
Lower Extreme Exoskeleton Rapid Terminal Sliding-Mode Robust
Control,” Journal of computers, vol.9, No.1,202-208. 2014.
\end{verbatim}

to reduce
sudden changes(command jump) at the phase transitions, a
smoothing method [5],[19] is introduced and a pre-transition
method is used to solve the swing delay due to the internal
pressure(approx 5 bar).
\begin{verbatim}
H. Kazerooni , Ryan Steger, Lihua Huang ”Hybrid Control of the
Berkeley Lower Extremity Exoskeleton (BLEEX),” Int. Journal of
Robotics Research, pp. 561-573, 2006.
oonbum Bae, Kyoungchul Kong, Masayoshi Tomizuka ”Gait Phase-
Based Control for a Rotary Series Elastic Actuator Assisting the knee
\end{verbatim}

The ground reaction
forces (GRF) magnitude and direction is used to command the
actuators. In some researches the GRF sensors are used
together with other sensors in the control architectures [17],
while in other researches the control system is based on the
GRF merely [RoboKnee, Honda].
\begin{verbatim}
17: K. Suzuki, G. Mito, H. Kawamoto, Y. Hasegawa, and Y. Sankai,
"Intention-based walking support for paraplegia patients with Robot
Suit HAL," Advanced Robotics, vol. 21, pp. 1441-1469, 2007
\end{verbatim}

\subsection{Survey literature and current exoskeleton systems for other approaches}

\subsection{How handle hand and other non-foot contact forces?}

Admittance control?

Map external contact forces to internal contact forces (operator-exoskeleton).

\section{Control Architectures}

A generic hierarchical control architecture~\cite{IEEE06907051}:

{\bf Lowest level Control:} present interface to actuators and
high bandwidth feedback control, degrees of freedom,
or synergies of either:
\begin{itemize}
\item
Position control.
\item
Pure force or torque control.
\item
Impedance control: specify desired position (for spring), and generate
stiffness, damping, and potentially a modified inertia or moment of inertia.
Modifying inertia is more dangerous than providing active stiffness
or damping.
Or (almost equivalently),
specify a desired force and change the force according to position, velocity,
and acceleration.
\end{itemize} 
One important synergy is coordinated control of the center of mass location and
velocity for balance.
PID or fancier feedback control laws are implemented at this level.
Gravity, friction, actuator dynamics, and other feedforward disturbance
compensation is implemented at this level, as well as any inverse dynamics
or other types of feedforward reference tracking control.
Disturbance observers for model errors and contact/wind forces
are implemented at this level.
Typically this is done by a high sampling rate global servo~\cite{Atlas-robot}
or local limb or joint-level high sampling rate servos~\cite{Sarcos-robot}.

>>>>

\title{A Survey of Possible Exoskeleton Control Architectures and
Algorithms\\
(Draft 2.0)}

@@

\section{Scope: What is this paper about?}

This paper surveys possible exoskeleton control architectures and
algorithms.
One goal is to address the question:
{\it How do exoskeleton control approaches compare with 
approaches to humanoid robot control?}
A companion paper surveys implemented exoskeleton control architectures
and algorithms~\cite{}.

The focus is on exoskeleton control that allows a
highly trained and top percentile athletic 
operator to carry a payload that weighs approximately the same amount
as the operator. We envisage these types of exoskeletons to be useful
in carrying protective and safety equipment for SWAT teams, police,
firefighters, and soldiers.

@@

We focus this survey on exoskeleton control for lower body tasks (standing, walking,
running, jumping, kicking, dodging, ...).
We do not survey exoskeleton control for human arms or manipulation.

We treat the torso and helmet of the exoskeleton as the payload,
and focus on a lower body exoskeleton to support any payloads that
are on the torso or head.

%\section{Symbiosis and Autonomy, Handling Errors, and Superhuman Reflexes}
A later white paper will discuss
possible combinations of response to user and exoskeleton
autonomy including balance control; response to trips, slips,
stumbles, and fumbles; task assistance and guidance (guide operator to
doorknob, button, or light switch); superhuman response to external
perturbations such as projectiles and explosions; and autonomous
execution (operator/exoskeleton symbiosis with multi-tasking)?

\subsection{Let's keep it simple}

A useful background paper: \url{http://summerschool.stiff-project.org/fileadmin/pdf/1804_C19.pdf}

Variants of nonlinear feedback control such as feedback linearization or
sliding mode control are largely ignored in this paper.
Once the decision to use feedback control
based on a set of observable quantities and with particular outputs is made,
one can try out the various linear and nonlinear feedback control paradigms
to see what works well.

Variants of function approximation methods such as lookup tables, fuzzy logic,
sigmoidal neural networks, radial basis functions, and locally weighted regression
are also largely ignored. Once the decision to use a function approximator
and what the inputs and outputs are has been made, one can try out the various
approaches to see what works well.

Same for optimization methods.

Same for constraint enforcement (such as avoiding self-collisions)
in either optimization or feedback control (barrier
Lyapunov functions (BLF)~\cite{IEEE06911561}).

Stability proofs of any of these methods should be viewed skeptically due
to the unmodeled operator-exoskeleton and exoskeleton-world contact dynamics, 
actuator unmodeled dynamics, joint play and exoskeleton
structural deformation, and controller
time delays which are typically ignored, especially in proofs involving
passivity arguments or Lyapunov functions.
We ignore stability proofs as well.

\section{What are some design philosophy alternatives?}

@@

Bypass valves and clutches could allow the operator to physically ``take over''
and push the relatively lightweight leg mounted exoskeleton around.

``Passive'' degrees of freedom can have physical or ``virtual'' springs
applied to assist in support (all abduct/adduct DOF, rotational DOF).
These aids can be clutched out or turned off when necessary.
``Virtual'' springs (springs implemented actively through exoskeleton
actuators) can be modulated.

@@

Can we build an exoskeleton that allows the operator/exoskeleton team
to be effective, but not necessarily behave naturally? 
The operator needs to learn to ``fly'' the suit,
just as operators learn to operate parachutes,
wing suits, diving equipment, high and low temperature protective
suits, firefighting equipment, and high altitude flight
and parachute suits.

@@

\section{Recommended Control Architecture}

We recommend the following hierarchical control architecture~\cite{IEEE06907051}:

{\bf Low Level Control:} Present interface to actuators and
high bandwidth feedback control, degrees of freedom,
or synergies of either:
\begin{itemize}
\item
Position control.
\item
Pure force or torque control.
\item
Impedance control: specify desired position (for spring), and generate
stiffness, damping, and potentially a modified inertia or moment of inertia.
Modifying inertia is more dangerous than providing active stiffness
or damping.
Or (almost equivalently),
specify a desired force and change the force according to position, velocity,
and acceleration.
\end{itemize} 
One important synergy is coordinated control of the center of mass location and
velocity for balance.
PID or fancier feedback control laws are implemented at this level.
Gravity, friction, actuator dynamics, and other feedforward disturbance
compensation are implemented at this level, as well as any inverse dynamics
or other types of feedforward reference tracking control.
Disturbance observers for model errors and contact/wind forces
are implemented at this level.
Typically this is done by a high sampling rate global servo~\cite{Atlas-robot}
or distibuted limb or joint-level high sampling rate servos~\cite{Sarcos-robot}.
Often, this controller is provided by the manufacturer of the robot
or exoskeleton and runs on a separate computer system from the
``user's'' or task/application controller~\cite{Sarcos,Atlas}.

{\bf ``Phoneme'' Control:} Mid-level controllers
for relatively constant phases of behavior.
Often involves position or velocity targets (goto or
stay-at X), trajectory references, ...
Example from walking: left-stance, double-support-1, right-stance, double-support-2, ...;
Example from running: left-foot-down, flight-1, right-foot-down, flight-2, ...;
Example from horizontal jumping: hjump-pushoff, hjump-flight, hjump-prepare-landing, hjump-impact, hjump-balance, ...;
Finer resolution may be needed. For example a walking stance period may
be broken up into heel-strike, first-half, second-half, pushoff, and toe-off.
The first and second halves of stance are divided by when the center of mass
passes over the ankle.
Local inverse kinematics (finding small joint motions for small target changes)
is performed at this level.
This level often provides the behavioral primitives or ``verbs'' for robot programming.

{\bf ``Word'' Control:} Mid-level controllers for sequences of ``phoneme'' behavioral
units.
Often involves finite state machines, timers, if-then condition testing, gain
scheduling, ...
Examples: walk-to-location( x=7.8m, y=2.3m ), run-to-location( speed=3.7m/s ),
horizontal-jump( height= 2m, length= 5m ).
Global inverse kinematics (selection of solution branch) is performed at this level.
This level often determines the control flow of a behavior.

{\bf Highest Level Control} aka {\bf Behavior Selection:} 
Answer the question: ``What do I do next?''
Select behaviors and behavioral parameters (targets,
speeds, durations).
Often involves finite state machines. timers, if-then condition testing, ...
This level often involves a great deal of human operator online interaction,
or is completely done by the operator.

{\bf Error Handling:} Errors must be detected and handled by parameter (target)
adjustment and/or behavior switching
at all of these levels.

This control architecture is intuitive, abstracts the hardware 
and lower level control in a straightforward way, is easy to program, and is
commonly used. Variants of this architecture were used by almost all teams
in the DARPA Robotics Challenge~\cite{}.

Closely related control architectures combine Phoneme and Word control,
or combine Phoneme, Word, and Behavior Selection.
This is not a major design change.

An alternative behavioral architecture might not have explicit levels but instead
has a ``soup'' of behaviors that compete or cooperate for the ability to
control the robot or exoskeleton. Subsumption architecture and other behavior-based
architectures are closer to the ``soup'' approach and to some extent
use inhibition and excitation
among behaviors to select active behaviors.

Another alternative behavioral architecture does not have explicit separate behaviors,
but one or a small number of functions or policies. 
We are seeing more of these architectures
due to the success and popularity of function approximation such as
deep learning.

\section{Exoskeleton dynamics}

Exoskeletons are similar to robots, but there are several important differences.
There are two sets of contact forces: operator-exoskeleton (which we denote
with a subscript $o$ as in $\vf_o$) and exoskeleton-world (which we denote
with a subscript $w$ as in $\vf_w$). If the operator is tightly strapped or
rigidly held by other types of physical constraints, we can lump the operator
body parts with the exoskelton parts which they are attached to in terms
of link inertias, location of center of mass, and moments of inertia,
and just think of the whole system as one robot.
We can also separate out the operator dynamics if this is more convenient
(we separate them out and ignore them in what follows).
If the operator is loosely connected to the exoskeleton, this introduces a
huge modeling and control problem, and would also be very dangerous for the
operator (as riding a bull or bucking horse in a rodeo is dangerous to a rider).

The dynamics of a revolute exoskeleton are:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \mJ_w^{\tr}(\vq) \vf_w = \vtau + \mJ_o^{\tr}(\vq) \vf_o
\label{eq:dynamics}
\end{equation}
where $\vq$ are the joint angles and position of the ``root'' of
the exoskeleton, $\dot{\vq}$ and $\ddot{\vq}$ are the corresponding
velocities and accelerations, $\mM$ is the inertia matrix, $\mC()$ are the Coriolis and
centripetal forces, $\mG()$ are the gravitational forces, $\mF()$ are the
friction forces, $\mJ_w^{\tr}(\vq) \vf_w$ (Jacobian matrix multiplied with the
contact forces)
are the exoskeleton-world
contact forces 
expressed as exoskeleton joint torques, 
$\vtau$ are the exoskeleton
joint torques,
and $\mJ_o^{\tr}(\vq) \vf_o$
are the operator-exoskelton forces expressed
as exoskeleton joint torques.
$\vtau$ is augmented with zeros corresponding to the dimensions of $\dot{\vq}$
that correspond to root velocities, since those dimensions are not actuated.

To understand these equations better, it is useful to have all Jacobians equal
to the identity matrix. This means the operator is directly applying torques
at the exoskeleton joints ($\vtau_o$) (a realistic assumption), 
and so is the world ($\vtau_w$) (very unrealistic assumption). In this case:
\begin{equation}
\mM(\vq) \ddot{\vq} + \mC(\vq,\dot{\vq}) + \mG(\vq) + \mF(\vq,\dot{\vq})
+ \vtau_w = \vtau + \vtau_o
\end{equation}

Note that we are ignoring the operator dynamics and operator muscle and tissue
forces resulting in operator joint torques. We will add these in later (or maybe
we won't ???).

We are also ignoring actuator dynamics, which we can handle by augmenting 
the state vector.

\section{Passive control with no active control}

One control option is to use only passive controls, and no active controls
or actuators.
Such a system would rely on counterbalancing, mechanical springs, air springs, etc.
to reduce forces. 
An operator would rely on mechanical backdrivability to drive
exokeleton to do tasks.
In this case, 
$\vtau$ is generated by passive devices such as springs or dampers.

\section{Open loop control}

Another control option is open loop control, in which actuator commands are
constant, or a function of time or phase of a behavior or task
(at least on the small time scale).
Commands or behaviors may be selected at a relatively coarse time interval
by the operator or by the exoskeleton controller.
This approach typically 
uses position and velocity control in either joint or
Cartesian coordinates
with either pre-generated references, selected targets,
or online generated trajectories to move to targets or perform tasks.
The operator relies on mechanical backdrivability 
to correct exoskeleton behavior.
In this case, it is common to generate the exoskeleton actuator commands using
a time or phase index:
$\vtau = \vtau_{ff}(t)$ where $\vtau_{ff}$ is a time
dependent feedforward torque vector.

Recognizing/estimating phase goes here.
Phase Triggered References could go here.
\begin{verbatim}
H.Kawamoto and Y.Sankai, “Power Assist Method Based on Phase
Sequence Driven by Interaction between Human and Robot Suit,”
Proceedings of the IEEE International Workshop on Robot and Human
Interactive Communication, pp. 491–496, 2004.
\end{verbatim}

@@

Central pattern generator

Iterative learning control to acquire $\vtau_{ff}(t)$ for desired trajectories.

\section{Active gravity compensation}

Active gravity compensation is similar 
to counterbalancing or physical gravity compensation, except it is done
with exoskeleton actuators to cancel the exoskeleton weight (but not inertia).
\begin{equation}
\vtau = \hat{\mG}(\hat{\vq})
\end{equation}
where $\vtau$ are the exoskeleton joint torques, 
and $\hat{\vq}$ is the estimated configuration of the exoskeleton,
The operator relies on mechanical backdrivability to drive
the exokeleton to do tasks.

Sarcos Primus Humanoid (ATR) video of active gravity compenstation applied
to a humanoid robot:\\
\url{https://www.youtube.com/watch?v=KNxxLm4sPys}

A simple approximation for a lower body exoskeleton with a heavy payload at the
torso is:
\begin{equation}
\vtau \approx \hat{\mJ}^{\tr}(\hat{\vq}) (\hat{m} \hat{\vg})
\end{equation}
where $\hat{\mJ}$ is the estimated Jacobian matrix,
$\hat{m}$ is the estimated
weight of the exoskeleton (which could include the operator),
and $\hat{\vg}$ is the gravity vector (it is estimated due to possible orientation
errors in the controller's estimate of vertical).

\section{Friction compensation}

Friction can also be compensated.
\begin{equation}
\vtau = \hat{\mF}(\hat{\vq},\hat{\dot{\vq}}) 
\end{equation}
are the estimated friction torques at the estimated exoskeleton configuration
and velocity.
Note that these may be load dependent, and may require additional state to
correctly handle stiction and hysteresis effects.

\section{Online inverse dynamics control}

The goal here is to map from desired motion (in this case desired
accelerations) to exoskeleton actuator commands (and desired operator
and world contact forces).
The exoskeleton actuator torques
attempt to match the torques caused by inertial, Coriolis, centripetal,
gravitational, frictional, and exoskeleton-world contact forces.
\begin{equation}
\vtau = \vtau_{\invdyn} 
= \hat{\mM}(\hat{\vq}) \ddot{\vq}_d
+ \hat{\mC}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mG}(\hat{\vq})
+ \hat{\mF}(\hat{\vq},\hat{\dot{\vq}})
+ \hat{\mJ}_w^{\tr}(\hat{\vq}) \hat{\vf}_w
\end{equation}
If there are actuator dynamics or other dynamics, we can include those
dynamics as well.

These force/torque estimates rely on estimated dynamic models and their parameters.
If the models and estimated state (position and velocity) are perfect, we get
perfect cancellation of most of the dynamics:
\begin{equation}
\mM(\vq) (\ddot{\vq} - \ddot{\vq}_d) = \mJ_o^{\tr}(\vq) \vf_o
\end{equation}
If $\ddot{\vq}_d$ is set to zero, the operator directly drives the acceleration
of the exoskeleton with an effective inertia of $(\mM^{-1}(\vq) \mJ_o^{\tr}(\vq))^{-1}$
\begin{equation}
\ddot{\vq} = \mM^{-1}(\vq) \mJ_o^{\tr}(\vq) \vf_o
\label{eq:cancel}
\end{equation}
$\ddot{\vq}_d$ can be used to drive the exoskeleton along trajectories, and
the operator can directly modify the trajectory.

@@

Online inverse dynamics control is a generalization of 
active gravity compensation to include
inertial, Coriolis, and centripetal forces (forces due to acceleration and
velocity), and potentially frictional forces.
Online inverse dynamics can be added to open loop behavior execution
and many other types of control such as operator force
feedback control, impedance control, and admittance control.
Computed torque control and feedback linearization are forms of online
inverse dynamics.

@@

\section{Operator-Exoskeleton Impedance Control}

@@

\section{Force feedback and get out of the way control (inverse dynamics version)}

@@

\section{Mapping External Forces To Operator Forces Using Force Control}

So far we have ignored or canceled the effect of 
the exoskeleton contact with the outside world.
A different control objective could be to map the external forces to 
desired operator forces:
\begin{equation}
{\vf_o}_d = F_{ow} ( \vf_w )
\end{equation}
and cancel the external forces not transferred in the inverse dynamics.
Everything in the above equation is estimated, so we are dropping the hats 
$\hat{\mbox{}}$.

For example, we could want the operator to feel a scaled (scale factor $\alpha$) 
version of the external
forces, as if they were contacting the operator directly.
\begin{equation}
{\vf_o}_d = \alpha J_o^{-\tr}(\vq) J_w^{\tr}(\vq) \vf_w
\end{equation}
and we scale the $\mJ_w^{\tr}(\vq) \vf_w$ term in the inverse dynamics
by $(1-\alpha)$
We are assuming we have any necessary force sensing between the exoskeleton and
the world.

If the operator applies joint torques, this simplifies to
\begin{equation}
{\vtau_o}_d = \alpha J_w^{\tr}(\vq) ( \vf_w )
\end{equation}
and the inversion of a Jacobian matrix is unnecessary.

This would provide a way of enabling the operator to perceive and control
external contact forces.

\section{Autonomous exoskeleton-world control}

\subsection{Autonomous exoskeleton-world position/velocity control}

A control objective could be to have the exoskeleton to move autonomously.

To be written.

\subsection{Autonomous exoskeleton-world force control}

A control objective could be to have the exoskeleton to apply a desired force.

To be written.

\subsection{Autonomous exoskeleton-world impedance control}

Another control objective would be to allow the exoskeleton-world contact to
be compliant or have an assigned impedance.

To be written.

\section{Using Online Optimization: Quadratic Programming}

The dominant paradigm in force controlled (typically
hydraulic) full size humanoid robots is to handle constraints in inverse dynamics
such as torque and center of pressure limits using online quadratic programming
to solve the inverse dynamics equations. This approach can also be used
to implement and trade off among the many control objectives described so far.

Quadratic programming~\cite{Wikipedia} solves the following problem:
Minimize
\begin{equation}
\vx^{\tr} \mQ \vx + c^{\tr} \vx
\end{equation}
with inequality constraints
\begin{equation}
\mA_1 \vx \le \vb_1
\end{equation}
and equality constraints
\begin{equation}
\mA_2 \vx = \vb_2
\end{equation}
The optimization variable $\vx$ includes accelerations $\ddot{\vq}$, joint
torques $\vtau$, and contact forces $\vf$.
For robots the discrepancy between actual and desired accelerations, joint
torques, and contact forces are all minimized:
\begin{equation}
C_r(\vx) = (\vq - \vq_d)^{\tr} \mW_{\vq} (\vq - \vq_d) + \vtau^{\tr} \mW_{\vtau} \vtau
+ \vf^{\tr} \mW_{\vf} \vf
\end{equation}
The dynamics (equation~\ref{eq:dynamics}) are enforced as an equality constraint.
Torque limits, friction limits, center of pressure location constraints,
and unidirectional contact constraints are all
enforced as inequality constraints.
Variants of this optimization
were performed every 1-2ms by various teams to control the
Atlas robots in the DARPA Robotics Challenge.

@@

\section{Operator-Exoskeleton Force Control using position controlled actuation}

@@

\subsection{Get out of the way control with position sources}

@@

\subsection{Operator force feedback with position sources}

@@

\section{Specific Proposed and Implmented Exoskeleton Control Schemes}

Sensitivity Amplification Control (SAC) used to control BLEEX
is a dynamic cancellation technique (a.k.a.
inverse dynamics, computed torque, or feedback linearization) that also tries
to adjust the apparent inertia of the exoskeleton (equation~\ref{eq:change-inertia1})
using acceleration feedback. It is not clearly stated but it appears that the
acceleration is the result of double differentiating position, as there do not
appear to be velocity sensors on BLEEX. Low pass filtering is applied to reduce
the high frequency noise amplified by the double differentiation process.

\subsection{Integral admittance control}

@@

@@

This approach implements a passive swing leg in addition to Sensitivity 
Amplification Control for the stance leg.

Stance: virtual joint torque control method Kazerooni et al.(2005)
When contact location between the wearer and the
exoskeleton is not fixed and difficult to estimate, this method
has been shown to be an effective method to generate the
locomotion for an exoskeleton robot.[3],[4],[21]
\begin{verbatim}
H. Kazerooni , Z. Racine , L. Huang and R. Steger ”On the control of
the Berkeley lower extremity exoskeleton (BLEEX),” Proc. IEEE Int.
Conf. Robot. Autom., pp. 4364-4371, 2005.
Racine Jean-Louis Charles ”Control of a Lower Extrmity Exoskeleton
for Human Performance Amplification,” University of California,2003.
582
Xiuxia Yang, Hongchao Zhao, Yi Zhang, Xiaowei Liu ”Carrying
Lower Extreme Exoskeleton Rapid Terminal Sliding-Mode Robust
Control,” Journal of computers, vol.9, No.1,202-208. 2014.
\end{verbatim}

To reduce
sudden changes(command jump) at the phase transitions, a
smoothing method [5],[19] is introduced and a pre-transition
method is used to solve the swing delay due to the internal
pressure(approx 5 bar).
\begin{verbatim}
H. Kazerooni , Ryan Steger, Lihua Huang ”Hybrid Control of the
Berkeley Lower Extremity Exoskeleton (BLEEX),” Int. Journal of
Robotics Research, pp. 561-573, 2006.
oonbum Bae, Kyoungchul Kong, Masayoshi Tomizuka ”Gait Phase-
Based Control for a Rotary Series Elastic Actuator Assisting the knee
\end{verbatim}

{\it
The ground reaction
forces (GRF) magnitude and direction is used to command the
actuators. In some researches the GRF sensors are used
together with other sensors in the control architectures [17],
while in other exoskeleton the control system is based on the
GRF merely [RoboKnee, Honda].
}
\begin{verbatim}
17: K. Suzuki, G. Mito, H. Kawamoto, Y. Hasegawa, and Y. Sankai,
"Intention-based walking support for paraplegia patients with Robot
Suit HAL," Advanced Robotics, vol. 21, pp. 1441-1469, 2007
\end{verbatim}

@@

\section{State Estimation}

We are surprised that acceleration feedback based on double differentiation
works well on real systems (even low pass filtered acceleration
estimated by a Kalman filter).
Impacts, shock waves, and the fact that neither
the operator's body parts or the exoskeletons parts are rigid bodies and the
joints are not well defined suggest this approach would be fragile to modeling
error and unmodeled dynamics.
It is not possible to distribute MEMs IMUs throughout the exoskeleton to
improve both velocity and 
acceleration estimation~\cite{Xinjilefu-thesis}, which should improve
Sensitivity Amplification Control.

\section{Discussion}

Discussion to be written.
